---
title: "Simple Linear Regression"
author: "Illarion Jabine"
date: "1/22/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Library
performance: Assessment of regression models performance.
Utilities for computing measures to assess model quality,
    which are not directly provided by R's 'base' or 'stats' packages.
    These include e.g. measures like r-squared, intraclass correlation
    coefficient, root mean squared error or functions to
    check models for overdispersion, singularity or zero-inflation and
    more. Functions apply to a large variety of regression models,
    including generalized linear models, mixed effects models and Bayesian
    models.

### Generic functions for fitted linear model objects

print() simple printed display
summary() standard regression output
coef() (or coefficients()) extracting the regression coefficients
residuals() (or resid()) extracting residuals
fitted() (or fitted.values()) extracting fitted values
anova() comparison of nested models
predict() predictions for new data
plot() diagnostic plots
confint() confidence intervals for the regression coefficients
deviance() residual sum of squares
vcov() (estimated) variance-covariance matrix
logLik() log-likelihood (assuming normally distributed errors)
AIC() information criteria including AIC, BIC/SBC (assuming normally distributed errors)

```{r}
library(performance)
lm_model <- lm(mpg ~ qsec, data = mtcars)

```

summary() method returns the “summary.lm” object, quite similar to those of “lm” objects, but now contain the summary information; the adjusted R2, the full matrix of coefficients, standard errors, t statistics, and p values.
```{r}
lm_model_summary <- summary(lm_model)
lm_model_summary$coefficients
lm_model_summary$fstatistic
```

### Checking model fit using check_model() from performance package


check_model() - Visual check of model various assumptions (normality of residuals, normality of random effects, linear relationship, homogeneity of variance, multicollinearity)
```{r}
check_model(lm_model)
```

