---
title: "1 sample t test"
author: "Illarion  Jabine"
date: "12/02/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```


### 1. Required packages:


* ggstatsplot: 
creates graphics with details from statistical tests included in the plots themselves. 
It provides an easier syntax to generate information-rich plots for statistical analysis of continuous (violin plots, scatterplots, histograms, dot plots, dot-and-whisker plots) or categorical (pie and bar charts) data. 
Currently, it supports the most common types of statistical approaches and tests: parametric, nonparametric, robust, and Bayesian versions of t-test/ANOVA, correlation analyses, contingency table analysis, meta-analysis, and regression analyses.

* ggpubr:
easy-to-use functions for creating and customizing 'ggplot2'- based publication ready plots.
* tidyverse
```{r load libraries, message=FALSE, warning=FALSE}
library(ggstatsplot)
library(ggpubr)
library(tidyverse)
```


### 2. Key terms and links
 * NHST
 https://libguides.library.kent.edu/SPSS/OneSampletTest
 http://www.sthda.com/english/wiki/one-sample-t-test-in-r
 https://yuzar-blog.netlify.app/posts/2021-12-20-one-sample-t-test-do-your-results-make-sense-or-how-to-compare-your-work-to-the-work-of-others/

### 3. The purpose of the test
One sample t test examines whether the mean of a population is statistically different from a known or hypothesized value. One sample t test is used to test statistical difference between a mean and a known or hypothesized value of the mean in the population. 
The test variable's mean is compared against a "test value", which is a known or hypothesized value of the mean in the population. 
Example: a machine is supposed to fill in bottles with 100 ml. A random sample is tested to check that the machine is not under- or over-filling the bottles.
One sample t test can only compare a single sample mean to a specified constant. It can not compare sample means between two or more groups. 

### 4. Check assumptions
Your data must meet the following requirements (from https://libguides.library.kent.edu/SPSS/OneSampletTest):
a. Test variable that is continuous (i.e., interval or ratio level)
b. Scores on the test variable are independent (i.e., independence of observations)
   * There is no relationship between scores on the test variable
   * Violation of this assumption will yield an inaccurate p value
c. Random sample of data from the population
d. Normal distribution (approximately) of the sample and population on the test variable
   * Non-normal population distributions, especially those that are thick-tailed or heavily skewed, considerably reduce the power of the test
   * Among moderate or large samples, a violation of normality may still yield accurate p values
e. Homogeneity of variances (i.e., variances approximately equal in both the sample and population)
f. No outliers
  1. Independence of observations (i.i.d.) - independent and identically distributed. The observations are not related (for example no autocorrelation)
  2. Homogeneity of variance: the variance within each group being compared is similar among all groups
  3. Check whether the data are normally distributed:
    3.1 Shapiro-Wilk normality test, where p-value > 0.05 would mean normally distributed data.
    Null hypothesis: the data are normally distributed
    Alternative hypothesis: the data are not normally distributed
Let's use iris data set for our one sample t test
```{r}
x <- unlist(subset(iris,Species == "virginica", select = Sepal.Width))
shapiro.test(x)

```
We fail to reject the null hypothesis.
    
    3.2 Quantile-Quantile plot, ggqqplot() function from ggpubr package. All points inside gray area mean normally distributed data.
```{r}
ggqqplot(x)

hist(x)
boxplot(x)
summary(x)
```
Our data is normally distributed, so the mean is a good representative for our sample. In this case we can use one sample t test. 
However if data are not normally distrubuted, then we choose a non-parametric one-sample Wilcoxon test which uses median as a test statistics.
 
### 5. Null hypothesis statistical test (NHST) procedure

Russell T. Warne suggest the following procedure:
1. Form groups in the data:
 There is a sample group and population.
 
2. Define the null hypothesis (H0). The null hypothesis is always that there is no difference between groups or that there is no relationship between independent and dependent variables:

For one sample t test:
H0: µ =  µ0  ("the population mean is equal to the [proposed] population mean")
H1: µ ≠  µ0  ("the population mean is not equal to the [proposed] population mean")
From the past research and after reviewing some articles on sepal width of virginica iris we set the null hypothesis that the mean length is 4 inches. 
So, we collected the samples to test H0.
```{r}
mu <- 4
```


3. Set alpha (α). The default alpha = .05.
```{r}
alfa <- 0.05
```

4. Choose a one-tailed or a two-tailed test. This determines the alternative hypothesis (H1).
From the prior research let us take two-tailed test.

5. Find the critical value, which is used to define the rejection region.
```{r}
t_critical <- qt(alfa/2,df = length(x)-1)
```

6. Calculate the observed value.
For one sample t test the observed value is calculated as follows:
 t = (mean(x)-mu)/(sd(x)/sqrt(n))
 mean(x) - Sample mean
 mu - The test value -- the proposed constant for the population mean
 (sd(x)/sqrt(n)) - Estimated standard error of the mean
 sd(x) - Sample standard deviation
```{r}
t_calc <- (mean(x)-mu)/(sd(x)/sqrt(length(x)))
```

7. Compare the observed value and the critical value. 
The calculated t value is then compared to the critical t value from the t distribution table with degrees of freedom df = n - 1 and chosen confidence level. If the calculated t value > critical t value, then we reject the null hypothesis. If the observed value is more extreme than the critical value, 
then the null hypothesis should be rejected. Otherwise, it should be retained.

We can see that t_calc is greater (in absolute terms) than t_critical.
Hence, we reject H0 that the mean length is 4 inches.

8. Calculate an effect size.

### 6. One sample t test in R
We can do one sample t test using t.test() base R:
```{r}
t.test(x,alternative = "two.sided",mu = mu)
```
Let us test another H0 that mu = 3:
```{r}
t.test(x,alternative = "two.sided",mu = 3)
```
We can not reject this H0.

We can use gghistostats() from ggstatsplot package:
```{r}
gghistostats(
    data       = iris %>% filter(Species == "virginica"),
    x          = Sepal.Width,
    test.value = mu,
    type       = "parametric", 
    normal.curve = TRUE
    )
```


